---
title: "GenPipes ATAC-seq Report"
subtitle: "Results from GenPipes 6.0.0 pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
params:
  project_id: NA
  primary_contact: NA
  project_start_date: NA
  readset_name: NA
  peak_annotations_tab: NA
---

```{css document style, echo=FALSE}
/* Make overflowing tables scrollable */
.scroll-box {
  overflow-x: auto;
}
/* Target table inside scroll-box */
.scroll-box table {
  border-collapse: separate;
  border-spacing: 0;
}
/* Freeze first column */
.scroll-box th:first-child,
.scroll-box td:first-child {
  position: sticky;
  background-color: white;
  left: 0;
  z-index: 1;
}

/* Format footer */
.footer {
  padding: 10px;
  text-align: center;
  background-color: #ebf3f5;
}

/* Format text block 1 (blue) */
.text-block-blue {
  background-color: #d5eef7;
  padding: 20px;
  border-radius: 5px;
}
/* Format text block 2 (gray) */
.text-block-gray {
  background-color: #ebf3f5;
  padding: 20px;
  border-radius: 5px;
}
  
/* Center images */
.center-image {
  display: block;
  margin-left: auto;
  margin-right: auto;
  margin-top: 10px;
  margin-bottom: 10px;
}
/* Format figure captions */
.figure-caption {
  padding: 5px;
  text-align: center;
  color: gray;
}

/* IGV floating images */
.igv-float-image {
  margin: 10px;
  border: 1px solid gray;
  border-radius: 5px;
}
/* IGV center images */
.igv-center-image {
  display: block;
  margin-left: auto;
  margin-right: auto;
  margin-top: 10px;
  margin-bottom: 10px;
  border: 1px solid gray;
  border-radius: 5px;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
  )

library(readr)
library(rmarkdown)
library(knitr)
library(ini)
library(kableExtra)
options(knitr.table.format = "html") 
library(stringr)
library(htmltools)
library(dplyr)
library(tibble)
library(data.table)
library(purrr)
library(rsvg)
#library(magick)
#library(pdftools)
```

```{r Finding images}
c3g_logo <- "./copied2report/report_images/c3g_logo.png"

igv_ref_genome1 <- "./copied2report/report_images/igv_ref_genome1.png"
igv_ref_genome2 <- "./copied2report/report_images/igv_ref_genome2.png"

igv_ref_genome_annotation1 <- "./copied2report/report_images/igv_ref_genome_annotation1.png"
igv_ref_genome_annotation2 <- "./copied2report/report_images/igv_ref_genome_annotation2.png"

igv_load_tracks <- "./copied2report/report_images/igv_load_tracks.png"

igv_group_autoscale <- "./copied2report/report_images/igv_group_autoscale.png"

igv_navigate <- "./copied2report/report_images/igv_navigate.png"
```

``` {r Finding files 1: ini file and readset}
pipeline_ini_file <- list.files(path="./copied2report", pattern="*.config.trace.ini",  all.files=TRUE, full.names=TRUE)
 
readset_file <- list.files(path="./copied2report", pattern=paste(params$readset_name), all.files=TRUE, full.names=TRUE)
```

``` {r Finding files 2: everything else}
#multiqc_html <- list.files(pattern="*.multiqc\\.html$", all.files=TRUE, full.names=TRUE) 
multiqc_html <- list.files(pattern="ChipSeq.chipseq.multiqc_data", all.files=TRUE, full.names=TRUE) #TMP until deeptools update

#multiqc_data <- list.files(pattern="*.multiqc_data$", all.files=TRUE, full.names=TRUE) 
multiqc_data <- list.files(pattern="ChipSeq.chipseq.multiqc.html", all.files=TRUE, full.names=TRUE) #TMP until deeptools update

qc_metrics_file <- list.files(pattern="^IHEC_.*_metrics_AllSamples\\.tsv$", all.files=TRUE, full.names=TRUE)

fp_plot_list <- list.files(path="./graphs/fingerprint_plots", pattern=".fingerprint.png", all.files=TRUE, full.names=TRUE)

insert_size_list <- list.files(path="./graphs/insert_size_plots", pattern="*.insert_size_histogram.pdf", all.files=TRUE, full.names=TRUE)
insert_size_example <- list.files(path="./copied2report", pattern="*.insert_size_histogram.png", all.files=TRUE, full.names=TRUE)

pca_png <- list.files(path="./graphs", pattern="read_coverage_pca.png", all.files=TRUE, full.names=TRUE)
pca_pdf <- list.files(path="./graphs", pattern="read_coverage_pca.pdf", all.files=TRUE, full.names=TRUE)

homer_html_file <- list.files(pattern="ChipSeq.homer_find_motifs_genome.md", all.files=TRUE, full.names=TRUE)

homer_known_list <- list.files(path="./annotation", pattern="knownResults.txt", all.files=TRUE, full.names=TRUE, recursive=TRUE)

sample1_motif1_logo_png <- "./copied2report/report_images/known1.logo.png" #For when logo png is created

tree_file <- list.files(pattern="tree.txt", all.files=TRUE, full.names=TRUE)
```

```{r C3G logo}
htmltools::img(src = knitr::image_uri(c3g_logo), 
               alt = 'C3G logo', 
               style = 'position:absolute; top:10px; right:250px; padding:20px',
               height = '125px')
```

```{r Define footer text}
footer_text <- "Canadian Centre for Computational Genomics • [Contact & Help Desk](https://computationalgenomics.ca/contact-us/)"
```

# {.tabset}

## Project & Pipeline Information

```{r Read ini file}
pipeline_ini <- read.ini(pipeline_ini_file)
```

### Project Details

<div class="text-block-blue">
**Project ID**: `r paste(params$project_id)`\

**Primary Contact**: `r paste(params$primary_contact)`\

**Project Manager**: François Lefebvre\

**Date of INI File Creation/Date Started**: `r paste(params$project_start_date)`\

**Assembly Genome Used**: `r paste(pipeline_ini$DEFAULT$assembly)` (`r paste(pipeline_ini$DEFAULT$assembly_synonyms)`)\

**Samples**: \

```{r Read readset file}
readset <- read.delim(readset_file, header = TRUE, sep = "\t")
```

```{r Samples table}
sample_df <- readset[,0:4]

sample_df_ordered <- sample_df[order(sample_df$Sample),] %>%
  mutate(MarkType = case_when(
          MarkType=="N" ~ "Narrow",
          MarkType=="B" ~ "Broad",
          MarkType=="I" ~ "Input"
          ))

kable(sample_df_ordered %>%
  select(Sample, Readset, MarkName, MarkType),
  col.names = c("Sample", "Readset", "Mark", "Mark Type"),
  row.names = FALSE) %>%
  kable_styling(bootstrap_options = "condensed", "responsive") %>%
  row_spec(0:nrow(sample_df_ordered), background = "white")
```
</div>

---

### Pipeline Details

ATAC-seq data is processed using GenPipes' ChIP-seq pipeline with the ATAC-seq option.

The table below summarizes the principal steps in pipeline.

For more details about GenPipes' ChIP-seq/ATAC-seq pipeline, visit the [Read The Docs](https://genpipes.readthedocs.io/en/genpipes-v-3.6.2/user_guide/pipelines/gp_chipseq.html) or [GitHub](https://github.com/c3g/GenPipes/tree/main/genpipes/pipelines/chipseq).

```{r Pipeline details 1: Find modules and module versions from ini file}
# Create list of key-value pairs from ini "DEFAULT" section
ini_default <- pipeline_ini[["DEFAULT"]]

# Initialize module_df
module_df <- data.frame(module = character(),
                        version = character(), 
                        link = character(),
                        embedded_link = character())

# Extract module names and versions from ini_default for module_df
for (i in names(ini_default)) {
  if (grepl("module", i, ignore.case = TRUE)) {
    module_name <- sub("mugqic/(.*?)/.*", "\\1", ini_default[i])
    module_version <- sub("([^/]*/){2}", "", ini_default[i])
    module_df <- add_row(module_df, module = module_name, version = module_version)
  }  
}
```


```{r Pipeline details 2: Add links to module_df}
# Links last checked: 2025-07-22
module_df <- module_df %>%
  mutate(link = case_when(
    module == "R_Bioconductor" ~ "https://www.bioconductor.org",
    module == "bwa" ~ "https://bio-bwa.sourceforge.net",
    module == "bwa-mem2" ~ "https://github.com/bwa-mem2/bwa-mem2/blob/master/README.md",
    module == "deepTools" ~ "https://deeptools.readthedocs.io/en/latest/",
    module == "GenomeAnalysisTK" ~ "https://gatk.broadinstitute.org/hc/en-us",
    module == "homer" ~ "http://homer.ucsd.edu/homer/motif/",
    module == "MACS2" ~ "https://github.com/macs3-project/MACS?tab=readme-ov-file",
    module == "MultiQC" ~ "https://seqera.io/multiqc/",
    module == "picard" ~ "https://broadinstitute.github.io/picard/",
    module == "samtools" ~ "https://www.htslib.org",
    module == "sambamba" ~ "https://lomereiter.github.io/sambamba/",
    module == "trimmomatic" ~ "http://www.usadellab.org/cms/?page=trimmomatic",
    module == "bedtools" ~ "https://bedtools.readthedocs.io/en/latest/",
    module == "star" ~ "https://pmc.ncbi.nlm.nih.gov/articles/PMC3530905/"
  ))

# Format as embedded links
module_df$embedded_link <- paste0("<a href='", module_df$link, "' target='_blank'>", module_df$module, "</a>")
```


```{r Pipeline details 3: Create pipeline_steps list}
## Define description and modules for each step of pipeline

# Descriptions for each step
qc_description = 
"Quality control checks are performed throughout the pipeline. The modules deepTools and Picard are used to evaluate various QC metrics and generate graphs. MultiQC aggregates all of the QC results into one report."

trimming_description = 
"Raw sequence reads are trimmed using Trimmomatic for quality and to remove sequencing adapters."

alignment_description = 
"Reads are aligned to a reference genome using the bwa-mem2 algorithm (from the Burrows-Wheeler Aligner (BWA) software package). Sambamba sorts reads according to their genomic coordinates and filters out duplicates, unmapped reads and low quality reads. Bedtools removes any reads from blacklist regions."

peak_calling_description = 
"Peaks are called with MACS2 to identify regions with enriched signals from the alignment data."

annotation_description = 
"The identified peaks are annotated with HOMER using the RefSeq annotations for the assembly genome used. HOMER also identifies motifs among the called peaks."

dba_description = 
"Sites with differential binding between sample groups are identified using the R Bioconductor package DiffBind."

# Extract module information from module_df for each step
qc_modules <- module_df %>% filter(module_df$module %in% c("deepTools", "MultiQC", "picard"))
trimming_modules <- module_df %>% filter(module_df$module == c("trimmomatic"))
alignment_modules <- module_df %>% filter(module_df$module %in% c("bwa", "bwa-mem2", "sambamba", "bedtools"))
peak_calling_modules <- module_df %>% filter(module_df$module == c("MACS2"))
annotation_modules <- module_df %>% filter(module_df$module == c("homer"))
dba_modules <- module_df %>% filter(module_df$module == c("R_Bioconductor"))

# Add the description and list of modules for each step
qc <- list("Description" = qc_description, "Module Information" = qc_modules)
trimming <- list("Description" = trimming_description, "Module Information" = trimming_modules)
alignment <- list("Description" = alignment_description, "Module Information" = alignment_modules)
peak_calling <- list("Description" = peak_calling_description, "Module Information" = peak_calling_modules)
annotation <- list("Description" = annotation_description, "Module Information" = annotation_modules)
dba <- list("Description" = dba_description, "Module Information" = dba_modules)

# Create a list of all steps
pipeline_steps <- list("Quality Control" = qc,
                       "Trimming" = trimming,
                       "Alignment" = alignment,
                       "Peak Calling" = peak_calling,
                       "Peak Annotation" = annotation,
                       "Differential Binding Analysis" = dba)
```


```{r Pipeline details 4: Flatten pipeline_steps list into df}
pipeline_df <- imap_dfr(pipeline_steps, function(step_data, step_name) {
  step_description <- step_data$Description
  modules <- step_data[["Module Information"]]
  if (is.null(modules) || nrow(modules) == 0) {
    tibble(
      Step = step_name,
      Description = step_description,
      Module = NA,
      Version = NA
    )
  } else {
    modules %>%
      mutate(
        Step = step_name,
        Description = step_description
      ) %>%
      select(Step, Description, Module = embedded_link, Version = version)
  }
})
```

```{r Pipeline details 5: Create table from pipeline_df}
kable(pipeline_df, "html", escape = FALSE, align = "clcc") %>%
  kable_styling(bootstrap_options = "basic") %>%
  column_spec(1, width = "125px") %>%
  column_spec(1:4, extra_css = "vertical-align: middle;") %>%
  collapse_rows(columns = c(1,2), valign = "middle")
```

---

### Citation
Mathieu Bourgey, Rola Dali, Robert Eveleigh, Kuang Chung Chen, Louis Letourneau, Joel Fillon, Marc Michaud, Maxime Caron, Johanna Sandoval, Francois Lefebvre, Gary Leveque, Eloi Mercier, David Bujold, Pascale Marquis, Patrick Tran Van, David Anderson de Lima Morais, Julien Tremblay, Xiaojian Shao, Edouard Henrion, Emmanuel Gonzalez, Pierre-Olivier Quirion, Bryan Caron, Guillaume Bourque, GenPipes: an open-source framework for distributed and scalable genomic analyses, GigaScience, Volume 8, Issue 6, June 2019, giz037, https://doi.org/10.1093/gigascience/giz037
\
\
**Please cite us and inform us when you do!**

Email: francois.lefebvre@mcgill.ca

Citation guide: [Read the Docs - Citing and Citations](https://genpipes.readthedocs.io/en/genpipes-v6.0.0/resources/citation.html#)

We like to keep track of published work.
 
---

<div class="footer">`r paste(footer_text)`</div> 

## Quality Control

### MultiQC Report

The MultiQC Report compiles quality control metrics for all samples assessed by various tools throughout the pipeline. It includes the following:

- General statistics 
- Trimmomatic: Surviving reads
- Samtools: Flagstat 
- Picard:
  - Alignment summary
  - Mean read length
  - Base distribution
  - Insert size
  - Mean base quality by cycle
  - Base quality distribution
- deepTools:
  - Correlation heatmap
  - Fingerprint plot
- HOMER:
  - Per sequence GC content
  - Tag length distribution
  - Chromosomal coverage
<!-- - MACS2: Filtered fragments -->

<div class="alert alert-success">
The MultiQC report can be found using the following link: **[MultiQC Report](`r paste(multiqc_html)`)**

It can also be found with the path ``r paste(multiqc_html)``
</div>

---

### QC Metrics

The table below presents several basic quality control metrics from ``r paste(basename(qc_metrics_file))``.

It contains information of about the number of raw reads for each sample along with the proportion of reads that were trimmed, mapped, duplicate, filtered, mitochondrial and singleton. 

It also includes the number of peaks identified for each sample, the number of mapped reads found within peaks and the Fraction of Reads in Peaks (FRiP) score. The FRiP score represents the number of reads in peaks over the total number of mapped reads; a greater FRiP score indicates a better signal-to-noise ratio. This is often used as measure of library quality. 

```{r Read IHEC metrics table}
qc_metrics <- read_tsv(qc_metrics_file)
qc_metrics_ordered <- qc_metrics[order(qc_metrics$Sample_Name),]
```

<div class="scroll-box">
```{r Create QC Metrics Table}
kable(qc_metrics_ordered %>%
  select(Sample_Name, Genome_Assembly, 
         Raw_Reads, Trimmed_Reads_Fraction,Mapped_Reads_Fraction, Duplicates_Reads_Fraction, Filtered_Reads_Fraction, 
         Mitochondrial_Reads_Fraction, Singletons_Reads,
         Nbr_Peaks, Reads_in_Peaks, frip)) %>%
  kable_styling(bootstrap_options = "hover", font_size = 12) %>%
  column_spec(1, width = "auto", extra_css = "white-space: nowrap;") %>%
  scroll_box(width = "100%", height = "100%")
```
</div>
\
<div class="alert alert-success">
 The full QC Metrics table can be found using the path ``r paste(qc_metrics_file)``
</div>

---

### Fingerprint Plots: Assessing Enrichment Signals

Fingerprint plots are used to assess the enrichment signals of samples and inputs.

Here, plots were generated using deepTools
[plotFingerprint](https://deeptools.readthedocs.io/en/develop/content/tools/plotFingerprint.html), a tool based on a method developed by
[Diaz *et al.*, 2014](https://github.com/songlab/chance/wiki/CHANCE-Manual#checking-the-strength-of-enrichment-in-the-ip).

For each sample, plotFingerprint randomly selects genomic regions as bins of a specified length and sums the per-base coverage within each bin. The bins are then ranked according to their coverage and the cumulative sum of the read coverage is plotted.

The rise of the cumulative read coverage curve indicates the specificity of the enrichment signals; while a steeper curve indicates more localized enrichment signals, a curve with a more gradual rise indicates more evenly distributed signals.

The type of curve expected depends on the type of mark is used:

- **Narrow peak marks** typically have steeply rising curves that indicate strong enrichment to specific regions.
- **Broad peak marks** typically have gradually rising curves that indicate more wide-spread enrichment.

#### Fingerprint plot example
```{r Get sample name for the fp plot example}
fp_plot1_sample <- sub("_.*", "", basename(fp_plot_list[1]))
```

Displayed below is an example of a fingerprint plot generated from your data for sample ``r paste(fp_plot1_sample)``.

This plot can be found with at ``r paste(fp_plot_list[1])``.
```{r Display Fingerprint plot example}
htmltools::img(src = knitr::image_uri(fp_plot_list[1]), 
               alt = "Fingerprint plot example", 
               class = "center-image")
```

#### Where to find fingerprint plots
<div class="alert alert-success">
Fingerprint plots for each samples can be found using the path ``r paste(dirname(fp_plot_list[1]))``

A fingerprint plot with all of the samples in one figure can be found in the *deepTools: Fingerprint plot* section of the **[MultiQC Report](`r paste(multiqc_html)`)**
</div>

---

### Insert Size Distribution Plots: Assessing Chromatin Accessibility

Insert size distribution plots display the frequency of DNA fragments by length, offering insight into chromatin accessibility patterns and serving as indicators of ATAC-seq library quality.

In high-quality ATAC-seq libraries, the insert size distribution typically exhibits decreasing periodic peaks corresponding to fragments within nucleosome-free regions (<100 bp) and fragments associated with mono-, di-, and tri-nucleosomes (~200, 400, and 600 bp, respectively). This pattern reflects the activity of the Tn5 transposase, which randomly tags and cuts accessible DNA within open chromatin regions, including linker regions between nucleosomes.

Insert size distribution plots were created using Picard's [CollectInsertSizeMetrics](https://gatk.broadinstitute.org/hc/en-us/articles/360037055772-CollectInsertSizeMetrics-Picard) tool. The red peaks correspond to the number of reads of a given insert size and the dashed line represents the cumulative fraction of reads below a given insert size.

#### Insert size distribution plot example
```{r Get sample name for the insert size plot example}
insert_plot1_sample <- sub("\\..*", "", basename(insert_size_example))
```

Displayed below is an example of a insert size distribution plot generated from your data for sample ``r paste(insert_plot1_sample)``.

This plot can be found with at ``r paste(dirname(insert_size_list[1]), sub("\\.png$", ".pdf", basename(insert_size_example)), sep="/")``.

```{r Display insert size distribution plot example}
htmltools::img(src = knitr::image_uri(insert_size_example),
               alt = "Insert size distribution example",
               class = "center-image")
```

#### Where to insert size distribution plots
<div class="alert alert-success">
Insert size distribution plots for each samples can be found using the path ``r paste(dirname(insert_size_list[1]))``

An insert size distribution plot with all of the samples in one figure can be found in the *Picard: Insert Size* section of the **[MultiQC Report](`r paste(multiqc_html)`)**
</div>

---

### Read Coverage PCA Plot: Comparing Samples

The amount of variability in read coverage between samples can be observed graphically with a Principal Component Analysis (PCA) plot.

This can useful for comparing the the variability between replicate samples and samples with different experimental conditions. Assessing the grouping patterns in a read coverage PCA plot can also be helpful for identifying potential particularities in your data, such as batch effects and outliers.

To assess genome-wide similarity between samples, deepTools [multiBamSummary](https://deeptools.readthedocs.io/en/develop/content/tools/multiBamSummary.html#multibamsummary) divides the genome into consecutive bins and calculates the read coverage for each bin for all samples. Then, deepTools [plotPCA](https://deeptools.readthedocs.io/en/develop/content/tools/plotPCA.html) evaluates the read coverage across bins between samples to generate the following two plots:

- A **PCA plot** with the eigenvalues of the top two principal components, representing the amount of variance explained by each principal component.
- A **Scree plot** with the amount of variance explained by each principal component (eigenvalue, blue bars), along with cumulative variance (red line) as a function of the top principal components.

Note that the read coverage PCA plot should only be used as a means of quality control and should not be used to to identify unknown groupings.

```{r Display PCA plot}
htmltools::img(src = knitr::image_uri(pca_png), 
               alt = "Read coverage PCA plot", 
               class = "center-image")
```

#### Where to find read coverage PCA plots
<div class="alert alert-success">
A PDF version of this read coverage PCA plot can be found using the path:
``r paste(pca_pdf)``

A PNG version can be found using the path:
``r paste(pca_png)``
</div>

---

<div class="footer">`r paste(footer_text)`</div> 

## Peak Annotation

### Motif Analysis with HOMER

DNA motifs are short recurring sequences of DNA that are usually thought to be associated with a biological function. 

Motif analysis of ATAC-seq data identifies transcription factor binding sites within accessible chromatin regions, offering insights into the regulatory elements and networks controlling gene expression.

It should also be noted that transcription factors actively bound to DNA can block Tn5 transposase cleavage within a binding-site, resulting in a depleted signal within the open chromatin region known as a "footprint".

[HOMER](http://homer.ucsd.edu/homer/) (Hypergeometric Optimization of Motif EnRichment) is a suite of tools that can be used for the analysis and annotation of next-generation sequencing data.

Here, HOMER was used to identify known motifs in the sequencing data.

#### Overview of HOMER's Known Motif Analysis Workflow

1. Peak sequence extraction: The DNA sequences corresponding to enriched regions are extracted based on the genomic coordinates of identified peaks.

2. Normalization: Target (sample) and background (input) sequence sets are compared and normalized according to their sequence composition to reduce bias.

3. Motif screening: Target sequence sets are screened for known motifs from the [HOMER Motif Database](http://homer.ucsd.edu/homer/motif/motifDatabase.html).

4. Output: HOMER generates an HTML report listing known motifs identified for each sample (see below), along with `.motif` files for each motif.

For more information on the HOMER motif analysis workflow, see [HOMER Motif Discovery Workflow](http://homer.ucsd.edu/homer/motif/motifFinding.html) (steps 1-4, 10-12, and 14 for screening known motifs).

---

### HOMER Identified Known Motif Example

```{r Read knownResults.txt file for the 1st sample}
# Read 1st file in list (1st sample)
known_results_sample1 <- read.delim(homer_known_list[1], header = TRUE, sep = "\t", dec = ".") 

# Get name of 1st sample
known_results_sample1_name <- str_match(paste(homer_known_list[1]), "annotation/([^/]+)/")[,2] 
```

Displayed below is a condensed example of a results table produced by HOMER's known motif analysis workflow. 

More specifically, the table shows the rank 1 motif identified by HOMER for sample ``r paste(known_results_sample1_name)``:

```{r Get logo of 1st motif of 1st sample}
# Get path to logos of 1st sample
known_results_sample1_path <- gsub(".txt", "", as.character(homer_known_list[1]))

# Path to logo of 1st motif
sample1_motif1_logo_svg <- paste(as.character(known_results_sample1_path), "known1.logo.svg", sep="/") 

# Convert sgv into png
rsvg_png(paste(sample1_motif1_logo_svg), paste(sample1_motif1_logo_png)) #sample1_motif1_logo_png defined in chunk Finding file 2

# Convert png to base64 URI and create html <img> tag
motif_logo_tag <- htmltools::img(
  src = knitr::image_uri(sample1_motif1_logo_png),
  alt = "Motif logo example"
  )
```

```{r Create sample 1 - motif 1 df}
# Create sample 1, motif 1 df
sample1_motif1_df <- known_results_sample1[1,] %>%
  select(Motif.Name, P.value, Log.P.value, q.value..Benjamini., X..of.Target.Sequences.with.Motif) %>%
  mutate(
    P.value = formatC(P.value, format = "e", digits = 2),
    q.value..Benjamini. = formatC(q.value..Benjamini., format = "e", digits = 2)
  )

# Add motif logo image to df
sample1_motif1_df$Logo.img <- as.character(motif_logo_tag)
```

```{r Display HOMER example table (1st motif from 1st sample)}
kable(sample1_motif1_df %>%
  select(Motif.Name, Logo.img, P.value, Log.P.value, q.value..Benjamini.,
         X..of.Target.Sequences.with.Motif),
  col.names = c("Name", "Motif", "p-value", "log p-value", "q-value (Benjamini)", "% of target sequence with motif"),
  escape = FALSE,
  align = "c"
  ) %>%
  kable_styling(bootstrap_options = "bordered", font_size = 12) %>%
  column_spec(c(3,5), width = "75px") %>%
  column_spec(4, width = "85px") %>%
  row_spec(0:1, extra_css = "vertical-align: middle;")
```

<div class="text-block-gray">
Each HOMER known motifs HTML report contains the following information:

- **Rank**: The relative rank of a motif according to its enrichment p-value
- **Motif**: The sequence logo showing the conservation of nucleotides in the motif
- **Name**: The name of the known motif in the [HOMER Motif Database](http://homer.ucsd.edu/homer/motif/motifDatabase.html)
  - Each motif name has the following structure:
\
    **`transcription factor name (DNA binding domain) / origin (GEO accession number) / generated by`**
\
    - **Transcription factor name**
    - **DNA binding domain**
    - **Origin**: The cell type and the immunoprecipitated protein used in the experiment where the motif was identified, along with its GEO accession number (otherwise its SRA accession number or the lead author’s name)
    - **Generated by**: The program/tool/resource used to generate the motif
- **P-value**: Enrichment p-value, where enrichment is determined by assessing how many target (sample) sequences are considered "bound" compared to the background (input)
- **Log P-value**: Enrichment log p-value
- **q-value (Benjamini)**: Enrichment q-value calculated using the Benjamini-Hochberg procedure
- **# of target sequences with motif**
- **% of target sequences with motif**
- **# of background sequences with motif**
- **% of background sequences with motif**
- **Motif File**: Links to the path of the `.motif` file
- **SVG**: Links to the path of the sequence logo image in SVG format
</div>

--- 

### HOMER Known Motif Analysis Results

The HTML pages with the known motifs for each sample can be found using the links below. 

<div class="alert alert-info">
```{r List of HOMER known motif links, results = 'asis'}
# ChipSeq.homer_find_motifs_genome.md has a list of html files for each sample:
homer_htmls <- read_lines(homer_html_file)
homer_htmls_ordered <- sort(homer_htmls)

for (html in homer_htmls_ordered) {
  if (grepl("Known", html)) {
    cat(html, sep = "\n") 
  }
}
```
</div>

<div class="alert alert-success">
Each of the HTML files corresponding to these links can also be found using the path format: `./annotation/SAMPLE/ATAC/knownResults.html`
</div>

---

<div class="footer">`r paste(footer_text)`</div>

## Genome Tracks 

### Visualizing Genome Tracks Using IGV

Integrative Genomics Viewer ([IGV](https://igv.org)) is a tool used for visually exploring genomic data by displaying sequencing data as tracks aligned to a reference genome.

For IGV's Quick Start Guide visit [Quick Start](https://igv.org/doc/desktop/#QuickStart/).

Summarized here are the steps to visualize your annotation data using the **IGV desktop application**.

<div class="text-block-blue">
#### 1. Launch IGV

If you don’t already have IGV installed on your desktop, visit the [Download IGV](https://igv.org/doc/desktop/#DownloadPage/) page to do so.
</div>
\
<div class="text-block-blue">
#### 2. Load a reference genome

A reference genome is used to map your data. It is important to select the same reference genome as the assembly genome used during the alignment step (find this in the *Project & Pipeline Information* tab).

Note: Because switching reference genomes clears loaded tracks, the reference genome should be loaded before the tracks.

There are 3 options for loading a reference genome:

  A. Selecting from the hosted genomes list\
  B. Loading a UCSC track hub\
  C. Defining a custom reference genome (not described here)\
  
```{r IGV reference genome image 1}
htmltools::img(src = knitr::image_uri(igv_ref_genome1), 
               alt = "IGV: Reference genome image 1", 
               width = "300",
               style = "float:right",
               class = "igv-float-image")
```
**A. Selecting from the hosted genomes list**

- In the menu bar, select ***Genomes***, then ***Download hosted genome...***
- A pop-up window will appear with a list of genomes hosted by IGV.
  - Select your reference genome.
  - If presented the option of downloading or using remote sequences and annotations, choose "Use remote".
  - Click ***OK***.
  
```{r IGV reference genome image 2} 
htmltools::img(src = knitr::image_uri(igv_ref_genome2), 
               alt = "IGV: Reference genome image 2",
               width = "500",
               class = "igv-center-image")
```
\

**Loading hosted annotation tracks (optional)**\
```{r IGV genome annotation images 1}
htmltools::img(src = knitr::image_uri(igv_ref_genome_annotation1), 
               alt = "IGV: Genome annotation image 1", 
               width = "400",
               style = "float:right",
               class = "igv-float-image")
```
Certain reference genomes also have related annotation tracks that can be loaded. 

To check what hosted tracks are available for your reference genome, select ***File*** then ***Load From Server*** in the menu bar.

```{r IGV genome annotation images 2}
htmltools::img(src = knitr::image_uri(igv_ref_genome_annotation2), 
               alt = "IGV: Genome annotation images 2",
               width = "850",
               class = "igv-center-image")
```
<div class="figure-caption">
Pictured here is an example of the hosted tracks available for the human (GRCh38/hg38) reference genome.
</div>

**B. Loading a UCSC track hub**\
UCSC hosts an archive of NCBI genome assemblies through [UCSC GenArk](https://hgdownload.soe.ucsc.edu/hubs/), which can be loaded into IGV.

- In the menu bar, select ***Genomes***, then ***Load Genome from UCSC GenArk…***
- Select the desired genome from the list and click ***Load***.
- The pop-up window "Select tracks to load" offers various annotation tracks related to the selected genome, including tracks from the following categories:
  - Mapping and sequencing
  - Variation and repeats
  - Genes and genes predictions
  - Expression and regulation
  - mRNA and EST
- After completing your selection of tracks, click ***OK***.

Visit IGV's User Guide: [Loading a reference genome](https://igv.org/doc/desktop/#UserGuide/reference_genome/%23select-a-hosted-genome) for more details.

</div>
\
<div class="text-block-blue">
#### 3. Load your data

Your track data is available in the form of bigWig files (.bw) that can be found within the `tracks` folder (expand `tracks.zip`).

- In the IGV menu bar, select ***File***, then select ***Load from File…***
- Find and select the relevant bigWig files and click ***Open***. It may take several minutes to load due to the size of the files.
- Once loaded, your tracks will appear in the track area above the reference genome.

```{r IGV load tracks image}
htmltools::img(src = knitr::image_uri(igv_load_tracks), 
               alt = "IGV: Load tracks image",
               width = "400",
               class = "igv-center-image")
```

Visit IGV's User Guide: [Loading and removing tracks](https://igv.org/doc/desktop/#UserGuide/loading_data/) for more details.
</div>
\
<div class="text-block-blue">
#### 4. Format your tracks

```{r IGV group autoscale image}
htmltools::img(src = knitr::image_uri(igv_group_autoscale), 
               alt = "IGV: Group autoscale image",
               width = "400",
               style = "float:right",
               class = "igv-float-image")
```
**Group autoscale**

Once all of your tracks are loaded, it is important to do a group autoscale of the tracks to make the peaks between them comparable. 

- Select all the tracks by holding the shift key and clicking each of the track titles on the left side of the page. The selected tracks will appear in grey.
- Right click anywhere in the grey area to open up the track options. Click ***Group Autoscale***.


**Other formatting options**

Other track formatting options are available by right clicking on any individual track or selection of multiple tracks. These include editing track labels and appearance (colour, display mode, etc.). 

</div>
\
<div class="text-block-blue">
#### 5. Navigate

Bring specific genomic regions into view by navigating through the genome tracks.

For more details on the various navigation controls, visit IGV's User Guide: [Navigating the view](https://igv.org/doc/desktop/#UserGuide/navigation/).

```{r IGV navigate image}
htmltools::img(src = knitr::image_uri(igv_navigate), 
               alt = "IGV: Navigate image",
               width = "800",
               class = "igv-center-image")
```
</div>
\
<div class="text-block-blue">
#### 6. Save

IGV sessions can be saved by clicking ***File*** > ***Save Session***.

IGV views can be saved as PNG or SVG images by clicking ***File*** > ***Save PNG image*** or ***Save SVG image***.

Visit IGV's User Guide on [Sessions](https://igv.org/doc/desktop/#UserGuide/sessions/) and [Saving Images](https://igv.org/doc/desktop/#UserGuide/saving_images/) for more details.

</div>

---

<div class="footer">`r paste0(footer_text)`</div>

## Output Explanation

### Output Explanation: `GenPipes_ATAC-seq_Results`

The folder `GenPipes_ATAC-seq_Results` contains files from the analysis of your ATAC-seq data using GenPipes. It includes all the files referenced in throughout this report. 

This page provides a summary and explanation of the different subfolders and files within `GenPipes_ATAC-seq_Results`.

The `GenPipes_ATAC-seq_Results` folder has the following structure:

```{r Output directory tree}
tree_lines <- readLines(tree_file)
tree_lines <- tree_lines[2:(length(tree_lines)-2)]

cat(
  paste0("Genpipes_ATAC-seq_Results", sep = "\n"), 
  paste0("├── ATAC-seq_Report.html", sep = "\n"),
  paste0(tree_lines, sep = "\n")
  )
```

### ``./annotation/``
<div class="text-block-blue">
The `annotation` folder contains files related to the ascription of biological meaning to your sequencing data. 

This folder contains a subfolder for each sample with the following:

- Peak annotation table `SAMPLE_ATAC.annotated.csv`
- Gene and genome ontology enrichment results (HTML files)
- Peak annotation results (as `.txt` files) from various tools and databases 
- HOMER motif enrichment results:
  - Known motifs results:
    - `knownResults.html`
    - `knownResults.txt`
    - `knownResults` folder, containing:
      - Sequence logos for each motif (SVG format)
      - `.motif` files for each motif
  - *De novo* results:
    - `homerResults.html`
    - `homerResults.txt`
    - `homerResults` folder, containing:
      - HTML files for each motif
      - Sequence logos for each motif (forward and reverse, SVG format)
      - `.motif` files for each motif
      - HTML files for similar *de novo* motifs

The `annotation` folder also includes a file called `peak_stats_AllSamples.csv` that compiles the peak statistics for all the samples in one.

For more information on the `.motif` files see [HOMER Motif Discovery Workflow](http://homer.ucsd.edu/homer/motif/motifFinding.html) step 12: Motif Files.

For more information on HOMER's *de novo* motif discovery methods see [HOMER Motif Discovery Workflow](http://homer.ucsd.edu/homer/motif/motifFinding.html) steps 1-9 and 12-13.
</div>

---

### ``r paste(multiqc_html)``
<div class="text-block-blue">
The file ``r paste(multiqc_html)`` leads to the [MultiQC Report](`r paste(multiqc_html)`) discussed in the *Quality Control* tab.
</div>

---

### ``r paste(qc_metrics_file)``
<div class="text-block-blue">
`IHEC_chipseq_metrics_AllSamples.tsv` contains various quality control parameters for each sample. 

A portion of ``r paste(qc_metrics_file)`` is displayed as a table in the *Quality Control* tab.
</div>

---

### `./graphs/`
<div class="text-block-blue">
The `graphs` folder contains the following:

- Miscellaneous graphs for each sample
  - Files: `SAMPLE.ATAC_Misc_Graphs.png` & `SAMPLE.ATAC_Misc_Graphs.ps`
  - Each file contains:
    - Location analysis of binding
    - Distribution of peaks found within exons
    - Distribution of peaks found within introns
    - Distribution of peak distances relative to TSS
- Quality control graphs for each sample and each input
  - Files for samples: `SAMPLE.ATAC_QC_Metrics.png`
  - Each file contains:
    - Tag count
    - Tag autocorrelation
    - Sequence bias
    - GC bias
- `fingerprint_plots/`
  - The folder `fingerprint_plots` contains PNG files of the fingerprint plots for each sample, an example of which can be found in the *Quality Control* tab. 
- `read_coverage_pca.pdf` & `read_coverage_pca.png`
  - Editable PDF and PNG versions of the read coverage PCA plots shown in the *Quality Control* tab. 

</div>

---

### `./tracks_bigWigs.zip` → `./tracks_bigWigs/`
<div class="text-block-blue">
`tracks_bigWigs.zip` contains annotation track files in bigWig format (.bw). These can be loaded into a genome browser such as [IGV](https://igv.org) or [USCS Genome Browser](https://genome.ucsc.edu) to visualize your sequencing data. 

Consult the *Genome Browser* tab of this report for a brief guide on visualizing your tracks in IGV.
</div>

---

### `trimReadsetTable.tsv`
<div class="text-block-blue">
`trimReadsetTable.tsv` contains statistics on the trimming process of the raw reads.

This file includes the number of raw reads for each sample as well as the number and percentage of surviving reads after the trimming step.
</div>

---

### Sharing this report and your results

To share this report and the results of the GenPipes processing of your ATAC-seq data, it is recommended to share `GenPipes_ATAC-seq_Results.tar.gz` as it contains all the files linked and referenced throughout the report.

---

<div class="footer">`r paste0(footer_text)`</div>
