---
title: "Peak Annotations Tab"
subtitle: "Child document to ChIP-seq_Report.Rmd"
---

## Peak Annotation

### Motif Analysis with HOMER

DNA motifs are short recurring sequences of DNA that are usually thought to be associated with a biological function. In ChIP-seq, identifying motifs can help determine sequence-specific binding sites for regulatory elements such as transcription factors.

[HOMER](http://homer.ucsd.edu/homer/) (Hypergeometric Optimization of Motif EnRichment) is a suite of tools that can be used for the analysis and annotation of next-generation sequencing data.

Here, HOMER was used to identify known motifs in the sequencing data.

#### Overview of HOMER's Known Motif Analysis Workflow

1. Peak sequence extraction: The DNA sequences corresponding to enriched regions are extracted based on the genomic coordinates of identified peaks.

2. Normalization: Target (sample) and background (input) sequence sets are compared and normalized according to their sequence composition to reduce bias.

3. Motif screening: Target sequence sets are screened for known motifs from the [HOMER Motif Database](http://homer.ucsd.edu/homer/motif/motifDatabase.html).

4. Output: HOMER generates an HTML report listing known motifs identified for each sample (see below), along with `.motif` files for each motif.

For more information on the HOMER motif analysis workflow, see [HOMER Motif Discovery Workflow](http://homer.ucsd.edu/homer/motif/motifFinding.html) (steps 1-4, 10-12, and 14 for screening known motifs).

---

### HOMER Identified Known Motif Example

```{r Read knownResults.txt file for the 1st sample}
# Read 1st file in list (1st sample)
known_results_sample1 <- read.delim(homer_known_list[1], header = TRUE, sep = "\t", dec = ".") 

# Get name of 1st sample
known_results_sample1_name <- str_match(paste(homer_known_list[1]), "annotation/([^/]+)/")[,2] 
# Get mark type of 1st sample
known_results_sample1_mark <- str_match(paste(homer_known_list[1]), paste0(known_results_sample1_name, "/([^/]+)/"))[,2]
```

Displayed below is a condensed example of a results table produced by HOMER's known motif analysis workflow. 

More specifically, the table shows the rank 1 motif identified by HOMER for sample ``r paste(known_results_sample1_name)`` with mark type ``r paste(known_results_sample1_mark)``:

```{r Get logo of 1st motif of 1st sample}
# Get path to logos of 1st sample
known_results_sample1_path <- gsub(".txt", "", as.character(homer_known_list[1]))

# Path to logo of 1st motif
sample1_motif1_logo_svg <- paste(as.character(known_results_sample1_path), "known1.logo.svg", sep="/") 

# Convert sgv into png
rsvg_png(paste(sample1_motif1_logo_svg), paste(sample1_motif1_logo_png)) #sample1_motif1_logo_png defined in chunk Finding file 2

# Convert png to base64 URI and create html <img> tag
motif_logo_tag <- htmltools::img(
  src = knitr::image_uri(sample1_motif1_logo_png),
  alt = "Motif logo example"
  )
```

```{r Create sample 1 - motif 1 df}
# Create smaple 1, motif 1 df
sample1_motif1_df <- known_results_sample1[1,] %>%
  select(Motif.Name, P.value, Log.P.value, q.value..Benjamini., X..of.Target.Sequences.with.Motif) %>%
  mutate(
    P.value = formatC(P.value, format = "e", digits = 2),
    q.value..Benjamini. = formatC(q.value..Benjamini., format = "e", digits = 2)
  )

# Add motif logo image to df
sample1_motif1_df$Logo.img <- as.character(motif_logo_tag)
```

```{r Display HOMER example table (1st motif from 1st sample)}
kable(sample1_motif1_df %>%
  select(Motif.Name, Logo.img, P.value, Log.P.value, q.value..Benjamini.,
         X..of.Target.Sequences.with.Motif),
  col.names = c("Name", "Motif", "p-value", "log p-value", "q-value (Benjamini)", "% of target sequence with motif"),
  escape = FALSE,
  align = "c"
  ) %>%
  kable_styling(bootstrap_options = "bordered", font_size = 12) %>%
  column_spec(c(3,5), width = "75px") %>%
  column_spec(4, width = "85px") %>%
  row_spec(0:1, extra_css = "vertical-align: middle;")
```

<div class="text-block-gray">
Each HOMER known motifs HTML report contains the following information:

- **Rank**: The relative rank of a motif according to its enrichment p-value
- **Motif**: The sequence logo showing the conservation of nucleotides in the motif
- **Name**: The name of the known motif in the [HOMER Motif Database](http://homer.ucsd.edu/homer/motif/motifDatabase.html)
  - Each motif name has the following structure:
\
    **`transcription factor name (DNA binding domain) / origin (GEO accession number) / generated by`**
\
    - **Transcription factor name**
    - **DNA binding domain**
    - **Origin**: The cell type and the immunoprecipitated protein used in the experiment where the motif was identified, along with its GEO accession number (otherwise its SRA accession number or the lead author’s name)
    - **Generated by**: The program/tool/resource used to generate the motif
- **P-value**: Enrichment p-value, where enrichment is determined by assessing how many target (sample) sequences are considered "bound" compared to the background (input)
- **Log P-value**: Enrichment log p-value
- **q-value (Benjamini)**: Enrichment q-value calculated using the Benjamini-Hochberg procedure
- **# of target sequences with motif**
- **% of target sequences with motif**
- **# of background sequences with motif**
- **% of background sequences with motif**
- **Motif File**: Links to the path of the `.motif` file
- **SVG**: Links to the path of the sequence logo image in SVG format
</div>

--- 

### HOMER Known Motif Analysis Results

The HTML pages with the known motifs for each sample can be found using the links below. 

Note that motif analysis is only performed on samples with a narrow peak mark type.

<div class="alert alert-info">
```{r List of HOMER known motif links, results = 'asis'}
# ChipSeq.homer_find_motifs_genome.md has a list of html files for each sample:
homer_htmls <- read_lines(homer_html_file)

for (mark in unique(sample_df$MarkName[sample_df$MarkType == "N"])) {
  cat("\n**Mark type: ", mark, "**\n", sep = "\n")
  for (homer_html in homer_htmls) {
    if (grepl(paste0("Mark ", mark), homer_html)) {
      if (grepl("Known", homer_html)) {
        cat(homer_html, sep = "\n") 
      }
    }
  }
}
```
</div>

<div class="alert alert-success">
Each of the HTML files corresponding to these links can also be found using the path format: `./annotation/SAMPLE/MARK/knownResults.html`
</div>

---

<div class="footer">`r paste(footer_text)`</div>
